Almost Orthogonal Vectors

### 2014-12-11 project3/paper/paper.tex
More changes

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$. (the above set must be non-empty since it must at least contain the element $N$)

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.

@@

Recall the Cauchy-Schwarz inequality for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

@@

@@

At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector. We present pseudocode for this algorithm above.

\begin{table}
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
    \hline
    \end{tabular}
    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our annealing method}
\end{table}

\section{A variation of the problem}

>>>>

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\},$$ where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$. (the above set must be non-empty since it must at least contain the element $N$)

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.

@@

Recall the Cauchy-Schwarz inequality for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right)\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right)\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

@@

\begin{table}
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
    \hline
    \end{tabular}
    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our numerical method}
\end{table}

At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector. We present pseudocode for this algorithm below.

As another note, in some instances the algorithm sometimes finds extrema that are not the extrema we have identified in cases where we provide optimal configurations. For example, for $n=3,N=4$, the $\epsilon$ value attained is very close to the optimal epsilon value of $\frac{1}{9}$, however the corresponding configuration is not a tetrahedron. This implies that in some cases there may be multiple configurations of vectors that are not equivalent up to rotation yet produce the optimal $\epsilon$ value.

@@

\section{A variation}
